Performance of two different ontological and data access methods

### 2014-07-01 cold2014/magnus.tex
trimming update abstract textit -> emph
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@703 43ef1190-90fd-4dd7-a769-d0121f02b80e

\begin{abstract}
  We evaluate two different approaches that have been used in the
  literature and in practice by different groups publishing metadata
  about cars when consumed by a prototype application. These
  approaches are characterized by a generic vs. domain specific
  ontology; by a constrained API vs. openly queryable data % TODO:
                                % Magnus, what else
  We have implemented a prototype application where a potential user
  may query selected properties of cars, and we investigate how the
  different choices made by the original designers influence the
  performance of the application. For the evaluation, we employ
  the statistical disipline of Design of Experiments.

@@

There will be several options on how to query the data because of all
the different specifications.  That is why we have chosen to use the
testing approach \textit{Design of Experiments} (DoE). This approach
will be further explained alongside the results in
Section~\ref{Results}. The evaluation will be based on four
experiments testing several aspects of the approaches. They will help
us determine what kind of factors are significant to the performance.

@@

% TODO differences between Daimler and Renault

@@

\subsection{Planning the experiment}
As other approaches, there are several steps in doing experiments with
DoE.  First one has to state an objective which is important to give
the experiment a purpose.  The next step is to choose a response. The
response is the experimental outcome or observation. There may be
multiple responses in an experiment. In this experiment there will be
one response and that is the response time on a HTTP post against the
application.  These first two steps are the same for all the
experiments in this thesis.

The next two steps of planning an experiment is specific to each
experiment itself. This means that these steps will be described more
in depth in each experiment because it may vary depending on the
actual experiment.  The third step in planning an experiment is
choosing factors and levels. This is a key part of DoE. A factor is a
variable that is studied in the experiment, and to be able to study
this factor it is needed to use two or more values of this
factor. These values are referred to as levels.  This varies from
benchmarking because often these levels are not required to do proper
benchmarking.  The levels will allow us the check each factor for its
significance. It is important to identify the key factors in the
planning stage. This is to get the maximum effect out of the
experiment.  Factors may be \textit{quantitative} or
\textit{qualitative}. Quantitative factors are often numerical values
that represent an interval. For instance the weight of a car is
considered a Quantitative value. Qualitative factors are predefined
values within a known set of values.  In our case a Qualitative factor
can be the type of fuel, for instance Diesel.

The fourth step in planning is to choose the experimental plan. Here
we will use one specific approach. The approach is called \textit{full
  factorial experiment} which we will go into more detail about in the
next sub-section. There are other approaches like fractional factorial
experiment, but they were not applicable or relevant to this thesis.
The last three steps of planning the experiment are performing the
experiment, analyzing the output and in the end drawing
conclusions. More about these steps will be taken individually at each
experiment.~\cite{PlanExp}

\subsection{Full factorial experiment}
A full factorial experiment is when one got $k$ factors and $n$ amount
of levels. This means that there is $n^k$ \textit{factorial designs}
which result in $n^k$ designs to execute.  In the thesis every
experiment will only have two levels, but the factors may vary. This
means that we will use a $2^k$ full factorial design for every
experiment. The experiments consist of $2^k$ combinations with $k$
factors. For instance if we have three factors where every factor got
two levels we get $2^3$ designs, also called \textit{runs}. In Table
\ref{designs} we can see all the possible runs we get from 3
factors. The + and - represent the level for each factor.  When
running the actual experiment one should use a \textit{planning
  matrix} to display the actual runs in the experiment. In Table
\ref{designsspec} we can see the same table as before, just that now
it is filled out with actual factors and their different levels.

\begin{table}
\begin{center}
    \begin{tabular}{ | l  l  l |}
    \hline
    {\bf Factor 1} & {\bf Factor 2} & {\bf Factor 3} \\ \hline
	 + & + & + \\ \hline
	  - & + & + \\ \hline
	 + & - & + \\ \hline
	  - & - & + \\ \hline
	 + & + & - \\ \hline
	  - & + & - \\ \hline
	 + & - & - \\ \hline
	  - & - &  - \\ \hline
    \end{tabular}
\end{center}
\caption{$2^3$ possible runs}\label{designs}
\end{table}

@@

Two key properties of full factorial designs are \textit{balance} and
\textit{orthogonality}.  Balance means that each factor level appears
the same amount of times in all the runs. For instance that the factor
Diesel appears the same amount of runs as Petrol. Orthogonal means
that two factors level combinations will appear in the same number of
runs. As an example will Diesel fuel combined with Automatic
transmission appear in the same number of runs as Petrol fuel with
manual transmission. We can see in Table \ref{designsspec} that both
these level combinations appear two times. Diesel and automatic
appears in run 1 and 5, while Petrol and manual appears in run 4 and
8.

Experiments like this can also be replicated, which just means that
the experimental plan is run more than one time.  It is not always
efficient to replicate an experiment. Sometimes it can be too
expensive to run it more than one time.  According to Wu and
Hamada(2009) replication is unnecessary for computer experiments
because repeated computer runs with the same input give the same
output.  If it is replicated the plan should be randomized to get the
best result. One reason for randomizing is to identify and get rid of
\textit{lurking variables}. A lurking variable is a factor that is not
taken into account when doing an experiment and that might affect the
outcome of the experiment. For instance if one is doing an experiment
outside. The outdoor temperature might be a affecting the experiment
if not taken into account.~\cite{FullFac}

\section{Experiments}
In this section we will present several experiments with different
amount of factors.  The objective is to find out how the ontologies
will perform in comparison and identifying which factors affect the
outcome of each experiment. We did the ontology alignment and
development of the application to prevent the loss of precision with
either ontology.

In the experiments the definition for VSO/COO is \textit{-vso} and for
CO is \textit{-co}.  These two definitions stand for which ontology
are tested in a particular run. In practice this means that when we
send in the keyword \textit{-vso} we only run the specification search
against the car models represented by VSO/COO, in our case the data
from Daimler. This is referred to as the ontology factor in each
experiment and is a factor in all the experiments which are presented
in this thesis. The ontology factor represents the difference in how
the data are represented and how the ontologies are structures, and
with this factor we want to determine how this affects the overall
performance.  We have programmed the application so that the querying
of the ontologies are done as similar as possible on both levels. This
was done to eliminate any interference from the application. There are
some differences in the programming due that the ontologies were
different and they will be discussed later on.  In Chapter
\ref{prototype}, the application and its differences are explained in
depth.

Before starting on the experiment we had to find a way to do the
actual experiment. We wanted to first create the planning matrix and
then execute each run as a HTTP post. To do this we needed to make a
small script to calculate the $2^k$ runs and perform the HTTP
posts. We chose to use the programming language \textit{python} to do
this. That was because python was a familiar language and it also had
a package for DoE\footnote{http://pythonhosted.org/pyDOE/}. With the
DoE package we just declared the amount of factors and the matrix was
created. At that point the matrix looked more like Table
\ref{designs}, which meant that we had to substitute the + and - with
the levels of each factor.  Further in the script we used the package
\textit{request}\footnote{http://requests.readthedocs.org/en/latest/}
to do the HTTP posts in python. Below one can see the line of code
executing the request.

@@

The \textit{payload} is a dictionary of all the values to post to the
application.  After the request we saved the response which contained
the response time of a run. After each request we saved the run plus
the response time and wrote it to file on CSV format. To evaluate the
result we used the programming language R with the packages
\textit{DoE.base}~\cite{DoEBase}, \textit{FrF2}~\cite{FrF2} and
\textit{BsMD}~\cite{BsMD}. R allowed us to parse the CSV files created
from the experiments and output them in two types of graphs.  A
similarity with all the experiments were that they all contained at
least the factors ontology, transmission and fuel type.

To evaluate the graphs we have used a method in the \textit{FrF2} R
package called \textit{DanielPlot}. This method assumes that the
estimated effects are normally distributed with means equal to the
effects. The means of all estimated effects are zero. Resulting in a
plot where the estimated effect would end up on a straight line. This
plot is then testing whether all the estimated effects have the same
distribution. All deviations of the straight line indicates a
significant factor. In our results we have used the absolute value of
the effects, also called a half-normal plot. The advantages of using
the half-normal plot is that all large estimated effects will end up
in the top right corner. This means that the highest performance
significance will end up in
the top right corner.~\cite{Plotting} \\
Here is a quick explanation of the axes shown in each graph.  The X
axis will show the absolute effect of each factor, shown in time. The
values on the X axis is represented in seconds which means that the a
plot will have a significance measured in seconds.  The Y axis shows
the half-normal scores. There are two way of representing the graph
and that is with either half-normal plots or normal plots. With
half-normal the effects on the X axis are shown with the absolute
effect. This means that all factors are placed with a positive
half-normal score. This is to avoid visual misleading scores because
with the normal scores, deviations from the line on negative scores
might confuse the reader who are evaluating the charts.  The scores
them selves are a measure for which factor that is the most
significant.  We have also set the \textit{alpha} to $0.05$ which
means that we will tolerate a false percentage of $5\%$.

@@

Like the rest of the experiments we used DanielPlot to evaluate and a
package in R to visualize the result. The Figure \vref{3factorGraph}
shows the results.  Here we see no significant factors. Still there
are small effects from several factors. In Table \vref{3factorEffect}
we can see the time effects from largest to smallest.  We see that the
ontology factor has the largest effect on approximately 2.2 seconds
which means that one of the ontologies overall is 2.2 seconds
faster. By reading the results from the linear model it is the factor
level \textit{-vso} which has an overall faster response time.

@@

\subsection{Experiment with ten factors}
The last experiment we did was a ten factor experiment. The goal here
was to see which factors would be deemed significant in a complete car
model search. With the results from this experiment we could discuss
several aspects of using the different ontologies.  Here we added 4
new factors in comparison to the six factor experiment. The newly
added factors were \textit{weight}, \textit{nr. of doors},
\textit{nr. of seats} and \textit{fuel consumption}. We used the same
approach here as in the previous experiments to calculate the average
value for the new factors. The average values can be seen in the
planning matrix in appendix \ref{appB}. With 10 factors the amount of
runs were $2^10$. This results in 1024 number of runs which made the
planning matrix too big to show here.

In the graph \vref{10factorGraph} we can see the results from running
all the 1024 runs.  We can see that there are a lot of significant
effects present here. There are more than any of the previous
experiments, but it is also more factors and factor interactions.
Unfortunately there are so many significant factors that there are
some of them that are illegible, but most of them have a low impact on
the significance graph. Like the six factor experiment we also see
here that there are three groups of significant factors. The first
group where most of them are placed range from around 2 seconds effect
to 11. Here there are a lot of factor interactions and why they are
there will be explained in the discussion.  The second group is very
like the second group we saw in the six factor experiment, the only
difference is that the total weight (WT) factor is substituted with
fuel consumption (FC). The last group is the same for all the
experiments, except the two factor experiment. Here we find the
ontology factor.  In Table \vref{10factorEffect} we can see the top
twenty absolute effects. This includes all the significant factors
from the last group and the second one, but also 5 from the first
group.  We can see that emission also here are present in a lot of the
entries, but so is ontology as well. The new thing here is the fuel
consumption has entered the most significant factors, and is by itself
higher than emission. The linear model again tells us that the option
\textit{-vso} has a significantly lower response time than
\textit{-co}

>>>>

\begin{abstract}
  We evaluate two different approaches that have been used in the
  literature and in practice by different groups publishing metadata
  about cars when consumed by a prototype application. These
  approaches are characterized by a generic vs. domain specific
  ontology; by a constrained tree-oriented API vs. openly queryable
  data; encouraging vs discouraging reasoning, etc.  We have
  implemented a prototype application where a potential user may query
  selected properties of cars, and we investigate how the different
  choices made by the original designers influence the performance of
  the application. For the evaluation, we employ the statistical
  disipline of Design of Experiments.

@@

There will be several options on how to query the data because of all
the different specifications.  That is why we have chosen to use the
testing approach \emph{Design of Experiments} (DoE). This approach
will be further explained alongside the results in
Section~\ref{Results}. The evaluation will be based on four
experiments testing several aspects of the approaches. They will help
us determine what kind of factors are significant to the performance.

@@

% TODO differences between Daimler and Renault
% TODO Alignment

@@

\subsection{Planning the experiment}
Like other approaches, there are several steps in doing experiments with
DoE.  First one has to state an objective which is important to give
the experiment a purpose.  The next step is to choose a response. The
response is the experimental outcome or observation. There may be
multiple responses in an experiment. In this experiment there will be
one response and that is the response time on a HTTP post against the
application.  These first two steps are the same for all the
experiments in this thesis.

The next two steps of planning an experiment is specific to each
experiment itself. This means that these steps will be described more
in depth in each experiment because it may vary depending on the
actual experiment.  The third step in planning an experiment is
choosing factors and levels. This is a key part of DoE. A factor is a
variable that is studied in the experiment, and to be able to study
this factor it is needed to use two or more values of this
factor. These values are referred to as levels.  The levels will allow
us to examine the influence of each factor. It is important to
identify the key factors in the planning stage. This is to get the
maximum effect out of the experiment.  Factors may be
\emph{quantitative} or \emph{qualitative}. Quantitative factors
are often numerical values that represent an interval. For instance
the weight of a car is considered a Quantitative value. Qualitative
factors are predefined values within a known set of values.  In our
case a Qualitative factor can be the type of fuel, for instance
Diesel.

The fourth step in planning is to choose the experimental plan. Here
we will use one specific approach. The approach is called \emph{full
  factorial experiment} which we will go into more detail about in the
next sub-section. There are other approaches like fractional factorial
experiment, but they were not applicable or relevant to this thesis.
The last three steps of planning the experiment are performing the
experiment, analyzing the output and in the end drawing
conclusions. More about these steps will be taken individually at each
experiment.~\cite{PlanExp}

\subsection{Full factorial experiment}
A full factorial experiment is when one got $k$ factors and $n$ amount
of levels. This means that there is $n^k$ \emph{factorial designs}
which result in $n^k$ designs to execute.  In the thesis every
experiment will only have two levels, but the factors may vary. This
means that we will use a $2^k$ full factorial design for every
experiment. The experiments consist of $2^k$ combinations with $k$
factors. For instance if we have three factors where every factor got
two levels we get $2^3$ designs, also called \emph{runs}. This results
in a \emph{planning matrix} that shows the actual runs in the
experiment as tabulated in Table~\ref{designsspec}. This contains the
chosen factors and their levels.

@@

\subsection{Running the experiments}
In this section we will present several experiments with different
amount of factors.  The objective is to find out how the ontologies
will perform in comparison and identifying which factors affect the
outcome of each experiment. We did the ontology alignment and
development of the application to prevent the loss of precision with
either ontology.

In the experiments the definition for VSO/COO is \emph{-vso} and for
CO is \emph{-co}.  These two definitions stand for which ontology
are tested in a particular run. In practice this means that when we
send in the keyword \emph{-vso} we only run the specification search
against the car models represented by VSO/COO, in our case the data
from Daimler. This is referred to as the ontology factor in each
experiment and is a factor in all the experiments which are presented
in this thesis. The ontology factor represents the difference in how
the data are represented and how the ontologies are structures, and
with this factor we want to determine how this affects the overall
performance.  We have programmed the application so that the querying
of the ontologies are done as similar as possible on both levels. This
was done to eliminate any interference from the application. There are
some differences in the programming due that the ontologies were
different and they will be discussed later on.  In Chapter
\ref{prototype}, the application and its differences are explained in
depth.

Before starting on the experiment we had to find a way to do the
actual experiment. We wanted to first create the planning matrix and
then execute each run as a HTTP post. To do this we needed to make a
small script to calculate the $2^k$ runs and perform the HTTP
posts. We chose to use the programming language \emph{python} to do
this. That was because python was a familiar language and it also had
a package for DoE\footnote{http://pythonhosted.org/pyDOE/}. With the
DoE package we just declared the amount of factors and the matrix was
created. At that point the matrix looked more like Table
\ref{designs}, which meant that we had to substitute the + and - with
the levels of each factor.  Further in the script we used the package
\emph{request}\footnote{http://requests.readthedocs.org/en/latest/}
to do the HTTP posts in python. Below one can see the line of code
executing the request.

@@

The \emph{payload} is a dictionary of all the values to post to the
application.  After the request we saved the response which contained
the response time of a run. After each request we saved the run plus
the response time and wrote it to file on CSV format. To evaluate the
result we used the programming language R with the packages
\emph{DoE.base}~\cite{DoEBase}, \emph{FrF2}~\cite{FrF2} and
\emph{BsMD}~\cite{BsMD}. R allowed us to parse the CSV files created
from the experiments and output them in two types of graphs.  A
similarity with all the experiments were that they all contained at
least the factors ontology, transmission and fuel type.

To evaluate the graphs we have used a method in the \emph{FrF2} R
package called \emph{DanielPlot}. This method assumes that the
estimated effects are normally distributed with means equal to the
effects. The means of all estimated effects are zero. Resulting in a
plot where the estimated effect would end up on a straight line. This
plot is then testing whether all the estimated effects have the same
distribution. All deviations of the straight line indicates a
significant factor. In our results we have used the absolute value of
the effects, also called a half-normal plot. The advantages of using
the half-normal plot is that all large estimated effects will end up
in the top right corner. This means that the highest performance
significance will end up in
the top right corner.~\cite{Plotting} \\
Here is a quick explanation of the axes shown in each graph.  The X
axis will show the absolute effect of each factor, shown in time. The
values on the X axis is represented in seconds which means that the a
plot will have a significance measured in seconds.  The Y axis shows
the half-normal scores. There are two way of representing the graph
and that is with either half-normal plots or normal plots. With
half-normal the effects on the X axis are shown with the absolute
effect. This means that all factors are placed with a positive
half-normal score. This is to avoid visual misleading scores because
with the normal scores, deviations from the line on negative scores
might confuse the reader who are evaluating the charts.  The scores
them selves are a measure for which factor that is the most
significant.  We have also set the \emph{alpha} to $0.05$ which
means that we will tolerate a false percentage of $5\%$.

@@

Like the rest of the experiments we used DanielPlot to evaluate and a
package in R to visualize the result. The Figure \vref{3factorGraph}
shows the results.  Here we see no significant factors. Still there
are small effects from several factors. In Table \vref{3factorEffect}
we can see the time effects from largest to smallest.  We see that the
ontology factor has the largest effect on approximately 2.2 seconds
which means that one of the ontologies overall is 2.2 seconds
faster. By reading the results from the linear model it is the factor
level \emph{-vso} which has an overall faster response time.

@@

\subsection{Experiment with ten factors}
The last experiment we did was a ten factor experiment. The goal here
was to see which factors would be deemed significant in a complete car
model search. With the results from this experiment we could discuss
several aspects of using the different ontologies.  Here we added 4
new factors in comparison to the six factor experiment. The newly
added factors were \emph{weight}, \emph{nr. of doors},
\emph{nr. of seats} and \emph{fuel consumption}. We used the same
approach here as in the previous experiments to calculate the average
value for the new factors. The average values can be seen in the
planning matrix in appendix \ref{appB}. With 10 factors the amount of
runs were $2^10$. This results in 1024 number of runs which made the
planning matrix too big to show here.

In the graph \vref{10factorGraph} we can see the results from running
all the 1024 runs.  We can see that there are a lot of significant
effects present here. There are more than any of the previous
experiments, but it is also more factors and factor interactions.
Unfortunately there are so many significant factors that there are
some of them that are illegible, but most of them have a low impact on
the significance graph. Like the six factor experiment we also see
here that there are three groups of significant factors. The first
group where most of them are placed range from around 2 seconds effect
to 11. Here there are a lot of factor interactions and why they are
there will be explained in the discussion.  The second group is very
like the second group we saw in the six factor experiment, the only
difference is that the total weight (WT) factor is substituted with
fuel consumption (FC). The last group is the same for all the
experiments, except the two factor experiment. Here we find the
ontology factor.  In Table \vref{10factorEffect} we can see the top
twenty absolute effects. This includes all the significant factors
from the last group and the second one, but also 5 from the first
group.  We can see that emission also here are present in a lot of the
entries, but so is ontology as well. The new thing here is the fuel
consumption has entered the most significant factors, and is by itself
higher than emission. The linear model again tells us that the option
\emph{-vso} has a significantly lower response time than
\emph{-co}
